Something about lineage metrics

### 2018-03-28 paper.tex
Update on Overleaf.

Evolution is an emergent effect of events that occur on a much finer-grained temporal scale. While this is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results. In computational evolutionary systems, we theoretically have adequate data to untangle these mechanisms. In practice, though, gathering appropriate data to do so can be challenging. Moreover, because of the vast difference between the scales that the causes and effects occur on, the quantity of such data can be overwhelming. Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population. Here, we will introduce a suite of such metrics and provide an intuition for what they can tell us about evolution.

\section{Methods}

\subsection{Metrics}

\subsection{Interpretation}

\subsubsection{Motivation}
Evolutionary biologists often accuse each other of making up "just-so" stories - post-hoc explanations for how an observed trait or pattern could have evolved. The problem with these stories is that, in biology, there is often no way to verify that they reflect what happened in biology. Due to evolution's stochastic nature, it is easy to come up with possible ways that something could have happened once, regardless of how repeatable they are. Theoretically, we should easily be able to avoid this problem in computational evolution by carefully checking the underlying assumptions behind our hypothesized explanation.  All too often, however, we fail to drill down to the true underlying mechanism behind an observed effect.

The metrics we propose here can help provide a check on this behavior, by making information about underlying evolutionary dynamics more readily accessible. They can only have that effect, however, if we completely understand what they are telling us about the way populations under different conditions are traversing different fitness landscapes. In order for these metrics to be useful for diagnosing the behavior of an evolving population, we need to establish ground truth for what underlying evolutionary dynamics are implied by different values of the metrics. For the purposes of building this intuition on a solid foundation, we wanted to be able to visualize the full evolutionary history of each population as it traversed the fitness landscape.

To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population. With these data, we were able to overlay successful lineages on top of the fitness landscape. Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition.

\subsubsection{Visualization methods}

@@

\subsection{Problems}

\section{Results and Discussion}

@@

>>>>

Evolution is an emergent effect of events (\textit{e.g.} replication, variation, and competition) that occur on a much finer-grained temporal scale. While this is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results. In computational evolutionary systems, we theoretically have adequate data to untangle these mechanisms. In practice, though, gathering appropriate data to do so can be challenging. Moreover, because of the vast difference between the scales that the causes and effects occur on, the quantity of such data can be overwhelming. Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population. Here, we  introduce a suite of such metrics and provide an intuition for what they can tell us about evolution.
% We test our intuitions on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths.

% Talk about abstract idea of a lineage being a sequence of states
% A lineage can be operationally defined as a sequence of states
% talk about different things those states could be

%  Talk about metrics being a useful concept for abstracting sequence of states (because doing stats on sequences of states is hard)

%Lit review - people have made pictures but it's hard to draw abstract large-scale inferences from them

% Introduce metrics

For the most part, the metrics we propose here will center around the evolutionary history of lineages in the population.

% @AML: Currently taking a crack at this bit of the intro (introducing lineages, etc) on paper at the moment!

% Something about: we often look at fitness over a lineage (i.e. fitness over time), we can look at other things...

\section{Metrics}

\subsection{Lineage Length}
\subsection{Mutation Accumulation} 
% Count/Magnitude (by type, aggregated etc
\subsection{Phenotypic Volatility}
% Useful things that this metric could identify:
% - Mutational bet hedging (h
% - Plasticity (high volatility)
% - Neutral landscapes (expect low volatility)

\subsection{Summary statistics}

\section{Test Problems}
To understand the metrics defined above, the test problems used need to be well understood and studied. The benchmark functions from [URL OR TITLE?] meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems. For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value. We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics.

\section{Data Analysis}

% Trim dramatically
%Evolutionary biologists often accuse each other of making up "just-so" stories - post-hoc explanations for how an observed trait or pattern could have evolved. The problem with these stories is that, in biology, there is often no way to verify that they reflect what happened in biology. Due to evolution's stochastic nature, it is easy to come up with possible ways that something could have happened once, regardless of how repeatable they are. Theoretically, we should easily be able to avoid this problem in computational evolution by carefully checking the underlying assumptions behind our hypothesized explanation.  All too often, however, we fail to drill down to the true underlying mechanism behind an observed effect.

%The metrics we propose here can help provide a check on this behavior, by making information about underlying evolutionary dynamics more readily accessible. They can only have that effect, however, if we completely understand what they are telling us about the way populations under different conditions are traversing different fitness landscapes. In order for these metrics to be useful for diagnosing the behavior of an evolving population, we need to establish ground truth for what underlying evolutionary dynamics are implied by different values of the metrics. For the purposes of building this intuition on a solid foundation, we wanted to be able to visualize the full evolutionary history of each population as it traversed the fitness landscape.

%To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population. With these data, we were able to overlay successful lineages on top of the fitness landscape. Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition.

@@

\section{Results and Discussion}
% Changing mutation: here's what happens when we change mutation on each problem
% - Metric 1
% Problem 1
% Problem 2
% Problem 3
% Problem 4
% - Metric 2
% ...

% Changing selection
% - Metric 1
% Problem 1
% Problem 2
% Problem 3
% Problem 4
% - Metric 2

@@

\section{Text Bank Because I'm not totally sure where things should go}
% Some gibberish on lineages
A lineage describes a chain of ancestors and their descendants where links in the chain are (typically) connected via parent-offspring relationships. In nature, every extant individual has a unique lineage that can be traced back to a single common ancestor. Depending on the modes of replication and variation employed along the lineage (\textit{e.g.} sexual reproduction, horizontal gene transfer), a lineage can be more or less challenging to disentangle. 
% something something something, we focus on asexual populations w/no HGT because easy starting place.

% Some gibberish on how we can represent lineages as sequences of states
Lineages can be imagined as a sequence of states where each state represents an individual. We can characterize each of these states: a phenotype, a genotype, a set of mutations (which may be empty) that differentiate genotype from parent genotype, time of birth/death, etc. 
Given sequences, can apply diagnostic metrics...

% Some gibberish on abstractions over the sequences of states
We can start to make abstractions: state = genotype, state = phenotype, state = location in environment.
Lineage length is measured over these abstractions:
%- Length in number of individuals 
%- Length in number of genotypes
%- Length in number of phenotypes
